---
title: "Practical Machine Learning"
author: "Peter Huber"
date: "Sunday, January 25, 2015"
output: html_document
---

Programming Assignment for Course "Practical Machine Learning", 01/2015, JHU on Coursera.

```{r setup-env, echo=FALSE, message=FALSE}
#loading libraries and setting seed, also set parallel option to enhance processing speed
library(caret)
library(doParallel)
registerDoParallel(cores=3)
```

```{r load-data, echo=FALSE}
#loading data; requires setwd to be set correctly
#setwd("<your path>")
data<-read.csv2(file="./pml-training.csv",header = T,sep = ",",dec=".")
test<-read.csv2(file="./pml-testing.csv",header = T,sep = ",",dec=".")
```

#Data cleaning

```{r clean-data, echo=FALSE, message=FALSE, cache=TRUE}
#data cleaning and preparation
doCleanData <- function(data) {
  dataInput<-data[data$new_window!="yes",]
  cleanedCols<-c()
  numRows<-dim(dataInput)[1]
  for (n in names(dataInput)) {
    sumNA <- sum( is.na(dataInput[,n]) | dataInput[,n]=="") 
    if((sumNA/numRows) < 0.1) {
      cleanedCols <- c(cleanedCols,n)
    }
  }
  cleanedCols <- cleanedCols[-grep(pattern=".*_timestamp.*",x=cleanedCols)]
  cleanedCols <- cleanedCols[-grep(pattern=".*_window$",x=cleanedCols)]
  cleanedCols <- cleanedCols[-which(cleanedCols %in% c("composedTime","user_name","X"))]
  dataInput<-dataInput[, cleanedCols]
}
cleanedData <- doCleanData(data)
````

The data cleaning process stripped away the majority of columns which do not contain sensor data.
The remaining columns are
```{r echo=FALSE, message=FALSE}
names(cleanedData)[order(names(cleanedData))]
```

#Model

From the data a training set of about 80% of is taken to build the model. the remaining 20% were
used to assses the model accuracy later on.
Random Forest (caret option "rf") was chosen as algorithm with a maximum of 7 predictors in 400 trees.
Preliminary attempts have shown that a value of more than 500 forest-trees or trees with more predictors than 5 do not add considerably more accuracy to the prediction. A optimization loop to find the perfect parameter set has not been included to keep this report short.

```{r build-model, echo=FALSE,message=FALSE,cache=TRUE,dependson="clean-data" }
#
#prepare training data; partition in train and test set
set.seed(2345)
inTrain <-  createDataPartition(cleanedData$classe, p=0.8, list = F)
trainData <- cleanedData[inTrain,]
testData <- cleanedData[-inTrain,]
#
#now train the model
ctrl <- trainControl(method="oob", repeats=2)
model <- train(classe~.,data=trainData,method = "rf",ntree=400,trControl=ctrl,tuneGrid = data.frame(.mtry = 7),importance=T)
```

#Analysis of model

To assess the model accuracy the Accuracy in and out of sample error has been computed.
Then a confusuion matrix gives an overview about the accuracy and miss-classifications class-wise
```{r echo=FALSE, message=FALSE}
##Predicting and analysis
##
##train set
predictions<-predict(model,newdata = trainData)
inSampleAccuracy<-sum(predictions==trainData$classe) / length(predictions)
##
##my test set
predictions<-predict(model,newdata = testData)
outOfSampleAccuracy<-sum(predictions==testData$classe) / length(predictions)
````
* Accuracy
```{r echo=FALSE, message=FALSE}
print(sprintf("Accuracy of predictions: In sample: %f; Out of sample: %f",inSampleAccuracy,outOfSampleAccuracy))
````
* Confusion Matrix
```{r echo=FALSE, message=FALSE}
confusionMatrix(reference = testData$classe, predictions)[[2]]
```` 
* Importance (MeanDecreaseInAccuracy)
```{r echo=FALSE, message=FALSE}
importance(model$finalModel)[,6]
````

#Prediction of test samples
```{r echo=FALSE, message=FALSE}
##
##jhu test set
jhuTestData <- doCleanData(test)
predictions<-predict(model,newdata = jhuTestData)
predictions
```
